An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-11-17 manuscript/main.tex
manuscript updates

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xurl}
\usepackage[colorinlistoftodos]{todonotes}

% specify link style
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=blue
}

@@

% override apa6 first paragraph index
% https://tex.stackexchange.com/questions/155028/indentation-apa6-class-for-2nd-paragraphs-after-title
% \makeatletter
%   \b@level@one@skip=-2.5ex plus -1ex minus -.2ex
%   \b@level@two@skip=-2.5ex plus -1ex minus -.2ex
% \makeatother

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Gabriele Chierchia$^3$, Sarah-Jayne Blakemore$^3$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA\\$^3$Department of Psychology, University of Cambridge, Downing Street, Cambridge, UK}

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit robust associations between item complexity and difficulty. However, we also find that item clones are not always psychometrically equivalent and cannot assumed to be exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials, results, and tutorials made available here will encourage researchers to use the MaRs-IB in their studies.}

\begin{document}
\maketitle

@@

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing geometric shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to a number of abstract rules. Based on these rules, participants must deduce which of four response options correctly completes the matrix. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of these elements across cells (Figure \ref{fig:fig00}a/b). To increase the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. A single item's clones share the same level of complexity (i.e. number of elements, number of rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses several of the issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB in which a sample of 659 adult, adolescent, and child participants had eight minutes to complete as many MaRs-IB items as possible. Of note, items were presented in the same order to all participants. The authors found that performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating satisfactory convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} made publicly available a summary of the difficulty (i.e. proportion correct responses) of each item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations in the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics, and therefore impede the design of new matrix reasoning tasks using the MaRs-IB. The first is that, given the fixed-order task design,  there was relatively few responses collected for the majority of the items. Indeed only 42 of 80 templates were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each template has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of how much information an item provides about ability. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test in measuring performance given a level of ability. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given researcher-specified constraints (e.g. test duration or difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning (i.e. difficulty, discrimination) of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory}, we specifically investigate how item complexity, i.e. the number of elements and number of rules, determine item functioning. We modeled these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. This explanatory analysis provides another means by which to validate the design of the MaRs-IB; that is, to determine whether the MaRs-IB exhibits the same associations between item properties and functioning as found in other established matrix reasoning tests. As a secondary benefit, explanatory IRT models yield more precise estimates of item parameters \citep{neuhaus2006separating}, which in turn help to ensure that any new MaRs-IB tests designed based on those estimates function as expected.

A third contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or if clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and discrimination can be generated simply by substituting a an item clone from one shape set for another. To test for exchangeability, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item difficulty and discrimination across clones, by distractor type and shape set, which must be negligible if clones are to be treated as exchangeable.

The remainder of the paper proceeds as follows. First we report a calibration study in which a large sample of adult participants completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that were derived in the calibration study.

@@

The purpose of the first study was to calibrate the items in the MaRs-IB using response data collected from a large number of adult participants recruited from the general population. In particular, we sought to accomplish the following three aims: (1) to quantify the psychometric properties (i.e. difficulty, discrimination) of the items in the MaRs-IB using item response models; (2) to measure the associations between item complexity and functioning; and (3) to determine whether item clones in the MaRs-IB are psychometrically equivalent and exchangeable.

@@

After providing consent, participants completed 16 items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3 x 3 matrix. Eight of the nine resulting cells contain abstract shapes, while one cell on the bottom right-hand side of the matrix is empty. Participants were instructed to ``complete the matrix'' by identifying the missing shape from among four possible alternatives.

@@

In order to ensure sufficient sampling of every item template and clone in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \citealt{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As such, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each clone by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or paired difference, PD; see \citealt{chierchia2019matrix} for details).

@@

We employed item response models to characterize the psychometric properties of each item. Specifically, we used additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In AMIS models, item parameters are defined according to a hierarchical structure in which item clones are nested in item templates. At the core of all of the AMIS models used here is the three parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item template $j$ is:

@@

In our analyses we considered two template (level 1) attributes: element number and rule number. These have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes among these shapes across cells. In the MaRs-IB there are four principal rules: size change, color change, position change, and shape change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Element and rule number are uncorrelated across item templates ($\rho$ = 0.176), and were determined for each item via manual annotation.

@@

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute. The allocation of geometric shapes to item clones in the MaRs-IB is complex, such that not all clones of the same shape set share the same perceptual elements (see the supplementary materials for a full discussion). In other words, shape set does not constitute a meaningful nominal variable and was therefore not included in the model. Instead, the contribution of perceptual elements to item functioning is captured by residual variability terms of the item structure models.

@@

\item Model 1: Item difficulty is a function of item template attributes (level-1) and item clone attributes (level-2). In this model, item difficulty is predicted solely by the fixed effect contribution of item attributes (i.e. element number, rule number, distractor type, response time).

\item Model 2: Item difficulty is a function of all item attributes (level-1/2) and level-1 residuals. The inclusion of the level-1 residual serves to quantify the magnitude of residual variability in item difficulty across item templates unexplained by the level-1 attributes (i.e. element number, rule number).

\item Model 3: Item difficulty is a function of all item attributes (level-1/2) and level-1/2 residuals. The inclusion of the level-2 residual serves to quantify the magnitude of residual variability in item difficulty across item clones unexplained by the level-2 attributes (i.e. distractor type, mean response time).

@@

In the second wave of model fitting, item discrimination was now a free parameter to be estimated. Moreover, in the second set of models, item difficulty parameters were specified according to the best-fitting model from the first wave. In the second wave, we fit an additional three models:

@@

    \item Model 4: Item discrimination is a function of item template attributes (level-1) and item clone attributes (level-2). In this model, item discrimination is predicted solely by the fixed effect contribution of item attributes (i.e. element number, rule number, distractor type, response time).

    \item Model 5: Item discrimination is a function of all item attributes (level-1/2) and level-1 residuals. The inclusion of the level-1 residual serves to quantify the magnitude of residual variability in item discrimination across item templates unexplained by the level-1 attributes (i.e. element number, rule number).

    \item Model 6: Item discrimination is a function of all item attributes (level-1/2) and level-1/2 residuals. The inclusion of the level-2 residual serves to quantify the magnitude of residual variability in item discrimination across item clones unexplained by the level-2 attributes (i.e. distractor type, mean response time).

@@

The AMIS modeling framework provides a natural means of accomplishing the three objectives of the calibration study. Via all of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses about the impact of these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can determine whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, a credible association between distractor type and either item parameter would entail that clones are not psychometrically equivalent. Moreover, if either level-2 residual variance term were non-negligible, that would similarly entail that not all item clones are equivalent due to unmodeled attributes (e.g. low-level perceptual features).

@@

All models were estimated within a Bayesian framework using Hamiltonian Monte Carlo as implemented in Stan (v2.22) \citep{carpenter2017stan}. For all models, four separate chains with randomised start values each took 7500 samples from the posterior. The first 5,000 samples from each chain were discarded. As such, 10,000 post-warmup samples from the joint posterior were retained. The $\hat{R}$ values for all parameters were less than 1.01, indicating acceptable convergence between chains, and there were no divergent transitions in any chain.

During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$. To ensure the identifiability of all models, person abilities were constrained to have a mean of zero and a variance of 1.0. Specifically, the residual variance of latent ability was specified as $V(\epsilon) = 1 - \sum \rho_p^2$.

@@

The goodness of fit of the models was compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. We computed the conditional leave-one-cluster out (LOCO) cross-validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held out items (rather than held out responses or held out participants) --- i.e., for inferring generalization to additional items constructed using the same features rather than to additional participants sampled from the same population.

@@

The $\chi^2_{NC}$ discrepancy statistic is a measure of item fit based on the comparison of observed and expected proportions of participants at each score level (here we ignore scores of zero and one, which comprised 7 total participants). A global measure of fit at the test level is obtained by summing the discrepancy values over the groups. In turn, SGDDM measures the mean absolute conditional correlation between all pairs of items; that is, the SGDDM is an index of residual inter-item correlations unexplained by the model. The SGDDM is important in this context because it tests for local dependence across items, the presence of which might otherwise lower the efficiency of future test forms by including nonindependent or redundant items. For both discrepancy statistics, the PPP value is the proportion of draws in which the posterior predictive discrepancy is equal to or higher than the realized discrepancy. A poor model fit to the data is indicated when the PPP-values are extreme (PPP $\leq$ 0.05 or PPP $\geq$ 0.95).

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 -- 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials (these were subsequently coded as incorrect responses). A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) solved all 16 of their items. Thus, over 95\% of observed scores were in a reasonable range. These results corroborate those reported by \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate in spite of the low-stakes testing environment.

Across all 384 item clones, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74; Figure \ref{fig:fig01}b). A total of 12 item clones (3.1\%) exhibited performance levels beneath chance, and only 22 item clones (5.7\%) had performance levels near ceiling ($\geq$ 90\%). Across items, the proportion of correct responses was negatively correlated with rule number ($\rho$ = -0.482, p < 0.001). These results further corroborate those reported by \cite{chierchia2019matrix} in that the MaRs-IB exhibited good functioning with performance levels for only a few item clones at ceiling or floor.

@@

There were also significant accuracy by difficulty ($\beta = 0.097$, $t = 14.34$, $p < 0.001$) and accuracy by rest score interactions ($\beta = -0.064$, $t = -8.76$, $p < 0.001$). Correct responses were even slower for more difficult items, but faster for better-performing participants. There was also a significant difficulty by score interaction ($\beta = 0.032$, $t = 6.55$, $p < 0.001$), such that better-performing participants exhibited even slower responses for more difficult items. This pattern of response time results is visible in Figure \ref{fig:fig01}c. The three-way interaction term was not significant.

>>>>

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}            % equation formatting
\usepackage{graphicx}           % figure formatting
\usepackage{xurl, hyperref}     % link formatting
\usepackage{xcolor,soul}        % highlight formatting

% specify link style
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    citecolor=blue
}

% define highlight style
\colorlet{usercolorname}{yellow!00}    % use for public version
% \colorlet{usercolorname}{yellow!30}    % use for reviewer version
\sethlcolor{usercolorname}

@@

% define title / short-title
\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}

% define authors
\author{Samuel Zorowitz$^1$, Gabriele Chierchia$^3$, Sarah-Jayne Blakemore$^3$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA\\$^3$Department of Psychology, University of Cambridge, Downing Street, Cambridge, UK}

% define abstract
\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the lack of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit robust associations between item complexity and difficulty. However, we also find that item clones are not always psychometrically equivalent and cannot assumed to be exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design two new sets of test forms that exhibited good score reliability and convergent validity with an established measure of matrix reasoning in an independent sample of adults (N=600). We hope that the materials, results, and tutorials made available here will encourage researchers to use the MaRs-IB in their studies.}

% start document
\begin{document}

\maketitle

@@

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing geometric shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to a number of abstract rules. Based on these patterns, participants must deduce which of four response options correctly completes the matrix. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of these elements across cells (Figure \ref{fig:fig00}a/b). Further increasing the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. An item's clones are equally complex (i.e. possess the same number of elements and rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses the copyright and limited reuse issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB in which a sample of 659 adult, adolescent, and child participants had eight minutes to complete as many MaRs-IB items as possible. Of note, items were presented in the same order to all participants. The authors found that measures of performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating satisfactory convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} made publicly available a summary of the difficulty (i.e. proportion correct responses) of every item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations of the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics and, consequently, impede the design of new matrix reasoning tests using the MaRs-IB. The first is that, given the fixed-order experiment design, there were relatively few responses collected for the majority of the items. Indeed only 42 of 80 item templates were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each item template has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of the reliability of an item. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test for measuring performance given a level of ability. Finally, IRT models make possible optimal test assembly \citep{van1998optimal}, or the design of new tests with maximal reliability given researcher-specified constraints (e.g. test duration or difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning (i.e. difficulty, discrimination) of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory}, we specifically investigate how item complexity, i.e. the number of elements and number of rules, determine item functioning. We studied these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. This explanatory analysis provides another means by which to validate the design of the MaRs-IB; that is, to determine whether the MaRs-IB exhibits the same associations between item properties and functioning as found in other established matrix reasoning tests. As a secondary benefit, explanatory IRT models yield more precise estimates of item parameters \citep{neuhaus2006separating}, which in turn help to ensure that any new MaRs-IB tests designed based on those estimates function as expected.

A third contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or if clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and reliability can be generated simply by substituting an item clone from one shape set for another. To test for exchangeability, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item difficulty and discrimination across clones, by distractor type and shape set, which must be negligible if clones are to be treated as exchangeable.

The remainder of the paper proceeds as follows. First, we report a calibration study in which a large sample of adult participants completed a number of items from the MaRs-IB. Using their response data, we fitted a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and item functioning. We conclude by reporting a second validation study, in which an independent sample of participants (N=600) completed one of five novel MaRs-IB test forms. These tests --- designed using optimal test assembly and the item parameter estimates from the first study --- were found to have good psychometric properties and convergent validity with an established measure of matrix reasoning ability. This second study provides a blueprint for how researchers can construct new MaRs-IB test measures based on the results of our first study.

@@

The purpose of the first study was to calibrate the items in the MaRs-IB using response data collected from a large number of adult participants from the general population. In particular, we sought to accomplish the following three aims: (1) to quantify the psychometric properties (i.e. difficulty, discrimination) of the items in the MaRs-IB using item response models; (2) to measure the associations between item complexity and functioning; and (3) to determine whether item clones in the MaRs-IB are psychometrically equivalent and exchangeable.

@@

After providing consent, participants completed 16 items from the MaRs-IB. The design of the MaRs-IB items have been described previously \citep{chierchia2019matrix}. Briefly, each MaRs-IB item consists of a 3x3 matrix. Eight of the nine cells contain abstract shapes, while one cell on the bottom right-hand side of the matrix is empty. Participants were instructed to ``complete the matrix'' by identifying the missing shape from among four possible alternatives.

@@

In order to ensure sufficient sampling of every item template and clone in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity, defined in \citealt{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As such, all participants received test forms of roughly equal difficulty.

Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each clone by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or paired difference, PD). \hl{Distractor type refers to the two strategies used to generate the distractor response options in the MaRs-IB. The MD strategy produces distractors that are variations of the target response. The MD strategy has the advantage of preventing pop-out effects (where the correct response ``pops out'' among the possible responses), but has the disadvantage of theoretically allowing participants to solve an item by looking at the response options only. In contrast, the PD strategy produces distractors that are have at least one component in common with the target response. The PD strategy prevents participants from solving items by looking at the response options alone, but can potentially induce pop-out effects. We note that, for complex items with many elements, it is not possible to prevent both at the same time.}

@@

We employed item response models to characterize the psychometric properties of each item. Specifically, we used additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. In AMIS models, item parameters are defined according to a hierarchical structure in which item clones are nested in item templates. The foundation of all the models used here is the three parameter logistic (3PL) item response model, where the probability of a correct response ($y_{ijk} = 1$) for person $i$ on item clone $k$ belonging to item template $j$ is:

@@

In our analyses we considered two template (level 1) attributes: element number and rule number. These have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes among these shapes across cells. In the MaRs-IB there are four principal rule types: size change, color change, position change, and shape change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Element and rule number are uncorrelated across item templates (Spearman's $\rho$ = 0.176), and were determined for each item via manual annotation.

@@

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute. The allocation of geometric shapes to item clones in the MaRs-IB is complex, such that not all clones of the same shape set share the same perceptual elements (see the supplementary materials for a full discussion). In other words, shape set does not constitute an interpretable categorical variable and was therefore not included in the model. Instead, the contribution of perceptual elements to item functioning is captured by the clone-level residual variability terms of the item structure models.

@@

    \item Model 1: Item difficulty is a function of item template attributes (level-1) and item clone attributes (level-2). In this model, item difficulty is predicted solely by item attributes (i.e. element number, rule number, distractor type, response time).

    \item Model 2: Item difficulty is a function of all item attributes (level-1/2) and level-1 residuals. The inclusion of the level-1 residual serves to quantify the magnitude of residual variability in item difficulty, across item templates, unexplained by the level-1 attributes (i.e. element number, rule number).

    \item Model 3: Item difficulty is a function of all item attributes (level-1/2) and level-1/2 residuals. The inclusion of the level-2 residual serves to quantify the magnitude of residual variability in item difficulty, across item clones, unexplained by the level-2 attributes (i.e. distractor type, mean response time).

@@

In the second wave of model fitting, item discrimination was specified as a free parameter to be estimated. Moreover, in the second set of models, the structure of the item difficulty parameters were specified according to the best-fitting model from the first wave. In the second wave, we fit an additional three models:

@@

    \item Model 4: Item discrimination is a function of item template attributes (level-1) and item clone attributes (level-2). In this model, item discrimination is predicted solely by item attributes (i.e. element number, rule number, distractor type, response time).

    \item Model 5: Item discrimination is a function of all item attributes (level-1/2) and level-1 residuals. The inclusion of the level-1 residual serves to quantify the magnitude of residual variability in item discrimination, across item templates, unexplained by the level-1 attributes (i.e. element number, rule number).

    \item Model 6: Item discrimination is a function of all item attributes (level-1/2) and level-1/2 residuals. The inclusion of the level-2 residual serves to quantify the magnitude of residual variability in item discrimination, across item clones, unexplained by the level-2 attributes (i.e. distractor type, mean response time).

@@

The AMIS modeling framework provides a natural means of accomplishing the three objectives of the calibration study. Via all of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses regarding the association between these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can determine whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, a credible association between distractor type and either item difficulty or discrimination would entail that clones are not psychometrically equivalent. Moreover, if we found any level-2 residual variance term were substantial, that would similarly indicate that not all item clones are equivalent due to unmodeled attributes (e.g. perceptual features).

@@

All models were estimated within a Bayesian framework using Hamiltonian Monte Carlo as implemented in Stan (v2.22; \citealt{carpenter2017stan}). For all models, four separate chains with randomised start values each took 7500 samples from the posterior. The first 5,000 samples from each chain were discarded. As such, 10,000 post-warmup samples from the joint posterior were retained. The $\hat{R}$ values for all parameters were less than 1.01, indicating acceptable convergence between chains, and there were no divergent transitions in any chain.

During estimation, the item discrimination parameters were restricted to be in the range $\alpha_{jk} \in [0, 5]$. To ensure the identifiability of all models, person abilities were constrained to have a mean of zero and a variance of 1.0. Specifically, the residual variance of latent ability was specified as $V(\epsilon) = 1 - \sum \rho_p^2$. The specification of the model priors are detailed in the supplement.

@@

The goodness of fit of the models was compared using Bayesian leave-one out cross-validation \citep{vehtari2017practical}, which has been found to perform better than more traditional information criteria for comparing item response models \citep{luo2017performances}. We computed the conditional leave-one-cluster out (LOCO) cross-validation \citep{merkle2019bayesian}, which measures a model's ability to generalize to held-out items (rather than held-out responses or held-out participants) --- i.e., for inferring generalization to additional items constructed using the same features rather than to additional participants sampled from the same population.

@@

The $\chi^2_{NC}$ discrepancy statistic is a measure of item fit based on the comparison of observed and expected proportions of participants at each score level (we ignore scores $\leq 1$, achieved by only 7 participants). A global measure of fit at the test level is obtained by summing the discrepancy values over the groups. In turn, SGDDM measures the mean absolute conditional correlation between all pairs of items; that is, the SGDDM is an index of residual inter-item correlations unexplained by the model. The SGDDM is important in this context because it tests for local dependence across items, the presence of which might otherwise lower the efficiency of future test forms via the inclusion of nonindependent or redundant items. For both discrepancy statistics, the PPP value is the proportion of draws in which the posterior predictive discrepancy is equal to or higher than the realized discrepancy. A poor model fit to the data is indicated when the PPP-values are extreme (PPP $\leq$ 0.05).

\hl{As a final validation of the best-fitting model, we also performed a parameter recovery analysis. In this analysis, we generated 100 artificial datasets with sampling and statistical properties matched to what we observed empirically. That is, we generated datasets with 1500 participants and 384 item clones (nested in 64 item templates), where the item parameters were randomly generated and matched to the observed distribution of item parameters. We then fit the best-fitting model to each artificial dataset, and quantified the consistency between the ground-truth and recovered model parameters. The complete details of this procedure can be found in the supplementary materials.}

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 -- 12.0; Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials (these were subsequently coded as incorrect responses). A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) solved all 16 of their items. Thus, over 95\% of observed scores were in a reasonable range. These results corroborate those reported by \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate in spite of the low-stakes testing environment.

Across all 384 item clones, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74; Figure \ref{fig:fig01}b). A total of 12 item clones (3.1\%) exhibited performance levels beneath chance, and only 22 item clones (5.7\%) had performance levels near ceiling ($\geq$ 90\%). Across items, the proportion of correct responses was negatively correlated with rule number ($\rho$ = -0.482, p < 0.001). These results further corroborate those reported by \cite{chierchia2019matrix} in that most items in the MaRs-IB exhibited good functioning, with performance at ceiling or floor for only a few item clones.

@@

There were also a significant accuracy by difficulty interaction ($\beta = 0.097$, $t = 14.34$, $p < 0.001$) and accuracy by rest score interaction ($\beta = -0.064$, $t = -8.76$, $p < 0.001$). Correct responses were even slower for more difficult items, but faster for better-performing participants. There was also a significant difficulty by score interaction ($\beta = 0.032$, $t = 6.55$, $p < 0.001$), such that better-performing participants exhibited even slower responses for more difficult items. This pattern of response time results is visible in Figure \ref{fig:fig01}c. The three-way interaction term was not significant.

In sum, we found evidence of a speed-accuracy trade-off in performance in the calibration sample. Both correct responses and better-performing participants were slower overall. Notably, better-performing participants demonstrated adjustments to their response times as function of item difficulty; whereas the lowest scoring participants maintained an equivalent work rate regardless of item difficulty, the highest scoring participants showed the largest slowing in responding as items became more challenging. We return to these results in the General Discussion.
